---
title: "650 Final"
format: 
  html:
    self-contained: true
editor: visual
---

# Read In Data and Libraries

```{r}
library(pacman)
p_load(tidyverse, lubridate, tidytuesdayR)

tuesdata <- tidytuesdayR::tt_load('2023-06-20')
ufo_sightings <- tuesdata$`ufo_sightings`
```

# Geography of UFO Sightings

## Countries

```{r}
# how many countries are in the set?
ufo_sightings |> 
  group_by(country_code) |> 
  summarize(
    count = n()
  ) |> 
# which countries have the most UFO sightings?
  arrange(desc(count))

# what proportion of all sightings happen in the US?
us_observations <- ufo_sightings |> 
  filter(country_code == "US")

nrow(us_observations)/nrow(ufo_sightings)

```

Given that the 91% of the UFO observations are form the United States, one of our research questions is: which states have the highest incidents of UFO sightings overall and per capita.

## States (Data Cleaning)

```{r}
# which states have the most UFO sightings overall
us_observations |> 
  group_by(state) |> 
  summarize(
    count = n()
  ) |> 
  arrange(desc(count))

# why are there 52 states?
unique_states <-us_observations |> 
  distinct(state)
print(unique_states)

# Fl is coded as both FL and Fl, need to code those observations as one unique state ID

count_fi <- us_observations %>%
  filter(state == "Fl") %>%
  nrow()

us_observations1 <- us_observations %>%
  mutate(state = ifelse(state == "Fl", "FL", state))

# check to see it worked
us_observations1 %>%
  filter(state == "Fl") %>%
  nrow()

# check number of US states now
unique_states1 <-us_observations1 |> 
  distinct(state)
print(unique_states) #51 makes sense because DC is included as a state

```

```{r}
#using cleaned ds # which states have the most UFO sightings overall
ufo_us <- us_observations1 |> 
  group_by(state) |> 
  summarize(
    count = n()
  ) |> 
  arrange(desc(count))

# Florida now has more overall observations than WA--highlighting the importance of working with clean data
```

## States UFO Observatiobns Per Capita

```{r}
# read in csv hosted on Github
state_pop_df <- readr::read_csv('https://raw.githubusercontent.com/gabrieldaiess/STAT650/main/us_pop_by_state.csv')

# join to UFO df

ufo_final_df <- inner_join(us_observations1, state_pop_df, by = c("state" = "state_code"))

# calculate per 100,000
ufo_final_df <- ufo_final_df |> 
  group_by(state) |> 
  mutate(pop_per_100k = (pop / 100000))

ufo_final_df <- ufo_final_df |> 
  group_by(state) |> 
  mutate(obs_per_100k = n() / pop_per_100k)

# group and rank observations per capita
ufo_pc <- ufo_final_df |> 
  group_by(state) |> 
  summarise(obs_per_100k = max(obs_per_100k, na.rm = TRUE)) |> 
  arrange(desc(obs_per_100k))
```

Per capita observations tells a much different story than overall count.

```{r}
# join needed for data viz
ufo_g1 <- inner_join(ufo_pc, ufo_us, by = c("state" = "state"))
```

## Figure 1

```{r}
# bar chart
options(repr.plot.width = 10, repr.plot.height = 15)  
ggplot(data = ufo_g1, aes(x = count, y = reorder(state, count), fill = obs_per_100k)) +
  geom_bar(stat = "identity", color = "black", size = 0.5) +
  scale_fill_gradient(low = "blue", high = "red") +
  labs(title = "UFO Sightings by State",
       x = "Count",
       y = "State",
       fill = "Obs. per 100K") +
  theme_minimal() +
  theme(axis.text.y = element_text(angle = -10, size = 5, face = "bold"))

```

# Shape:

```{r}
#count shapes of UFOs
shape_count <- ufo_sightings |> 
  count(shape) |>
  arrange(desc(n))
shape_count
```

## Figure 2

```{r}
# make a figure of shape
ggplot(data = shape_count, aes(x = n, y = reorder(shape, -n))) + 
  geom_bar(stat = "identity", fill = "orange") + 
  labs(title = "UFO Sightings by Shape",
       x = "Shape",
       y = "Sightings Count") +
  theme_minimal()
```

# Duration

How long do most UFO sightings last?

The UFO sighting durations, as shown in the histogram, are right-skewed, indicating most sightings are of shorter durations, particularly below 10 minutes. The highest frequency falls within the 10-100 minutes range. While there are infrequent cases of prolonged sightings lasting beyond 10,000 minutes, these are rare. The log scale on the x-axis aids in visually understanding this broad range of durations by emphasizing the multiplicative differences, making the distribution more comprehensible.

```{r}
#summary states on duration for paper
ufo_sightings <- ufo_sightings |>
  mutate(duration_minutes = duration_seconds / 60)


summary_stats <- ufo_sightings |>
  summarize(
    mean_duration = mean(duration_minutes, na.rm = TRUE),
    median_duration = median(duration_minutes, na.rm = TRUE),
    max_duration = max(duration_minutes, na.rm = TRUE),
    min_duration = min(duration_minutes, na.rm = TRUE)
  )
  summary_stats
```

## Figure 3

```{r}
ggplot(ufo_sightings, aes(x = log10(duration_minutes))) +
  geom_histogram(fill = "deeppink", color = "black", binwidth = 0.5) +
  labs(
    title = "Histogram of UFO Sighting Durations",
    x = "Log 10 (Duration)",
    y = "Frequency"
  ) +
  theme_minimal()
```

# Hour/Month Sightings Occur

## Figure 4

```{r}
ufo_sightings <- ufo_sightings |>
  mutate(hour = hour(reported_date_time))

ggplot(ufo_sightings, aes(x = hour)) +
  geom_bar(fill = "cyan", color = "black") +
  labs(title = "UFO Sightings by Hour of the Day", x = "Hour of the Day", y = "Number of Sightings") +
  theme_minimal()


```

```{r}
ufo_sightings <- ufo_sightings |> 
  mutate(
    month = month(reported_date_time, label = TRUE),
    year = year(reported_date_time),
    # Create a decade variable
    decade = 10 * (year(reported_date_time) %/% 10)
  )
```

## Decade Analysis

```{r}
#Determine the current decade
current_decade <- 10 * (year(Sys.Date()) %/% 10)

#Filter for the last 5 decades
ufo_sightings <- ufo_sightings |> 
  filter(decade >= current_decade - 40) 

```

## Summarizing sightings by decade and month

```{r}
decade_sightings <- ufo_sightings |>
  group_by(decade, month) |>
  summarize(count = n(), .groups = "drop")
```

## Figure 5

```{r}
ggplot(decade_sightings, aes(x = month, y = count, group = as.factor(decade))) +
  geom_line(aes(color = as.factor(decade))) + 
  labs(
    title = "UFO Sightings by Month (Last 5 Decades)",
    x = "Month",
    y = "Number of Sightings",
    color = "Decade"
  ) +
  theme_minimal()
```
